---
title: "PS270 Final Project"
author: "Matt Peters"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Milestone 1

I am interested in exploring data to help generate insights into the effectiveness of educational programs, including classroom practices, standards for teacher education, assessments, and curricula, in order to improve learning outcomes. I might just want to focus on Wisconsin, or alternatively have a comparative approach between states. 

## Milestone 2

For my research, I plan on utilizing data from the Wisconsin Department of Public Instruction to examine the relationship between the percentage of students receiving free or reduced-price lunch and student attendance rates in Wisconsin public schools. My research question is: How does the socioeconomic status of students, as measured by free or reduced-price lunch eligibility, affect their attendance rates? I hypothesize that higher percentages of students eligible for free or reduced-price lunch will correlate with lower attendance rates. I am basing this on the theory that socioeconomic challenges can hinder a student's ability to attend school regularly, whether that be originating from financial instability, lack of transportation, health issues, or more. The independent variable, percentage of students receiving free or reduced-price lunch, is measured as a percentage of the student population and varies across different school districts. The dependent variable, student attendance rates, is measured as the percentage of students attending school regularly. A negative correlation, where higher percentages of students receiving lunch assistance are associated with lower attendance rates, would support my hypothesis. Conversely, if the data shows no significant correlation or a positive relationship, it would challenge my hypothesis, indicating that other factors, such as parental support or school culture, may play a more significant role in student attendance.

## Milestone 3

```{r}
library(tidyverse)
library(readxl)
```

### Reading and combining data sets for Wisconsin school districts

```{r}
meals <- read_xlsx("data/eligibility_district_oct_2022-23.xlsx")
attendance <- read_csv("data/attendance/attendance_certified_2022-23.csv")
```

```{r}
meals <- meals |> 
  mutate(District_Code = str_sub(`Agency Code`, -4))

combined <- left_join(meals, attendance, by = c("District_Code" = "DISTRICT_CODE"))
```

```{r}
combined_all <- combined |> 
  filter(GROUP_BY_VALUE == "All Students",
         GRADE_GROUP == "[All]") |> 
  select(-"Independent Charter") |> 
  select(-(District_Code:STUDENT_COUNT)) |> 
  mutate(POSSIBLE_DAYS_OF_ATTENDANCE = as.numeric(POSSIBLE_DAYS_OF_ATTENDANCE),
         ACTUAL_DAYS_OF_ATTENDANCE = as.numeric(ACTUAL_DAYS_OF_ATTENDANCE),
         ATTENDANCE_RATE = as.numeric(ATTENDANCE_RATE),
         ATTENDANCE_RATE = ATTENDANCE_RATE / 100)
```

```{r}
ggplot(combined_all, aes(x = `% Free and Reduced`, 
                         y = ATTENDANCE_RATE)) +
  geom_point(alpha = 0.5, color = "steelblue3") +  
  geom_smooth(method = "lm", se = FALSE, color = "indianred2") +
  labs(title = "Relationship Between Free/Reduced Lunch Eligibility and\nAttendance Rates in Wisconsin Public School Districts (2023)",
       x = "% of Students Receiving Free/Reduced Lunch by School District",
       y = "Attendance Rate by School District (%)",
       caption = "Source: Wisconsin Department of Public Instruction") +
  scale_x_continuous(labels = scales::percent) + 
  scale_y_continuous(labels = scales::percent) + 
  theme_light() +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## Milestone 4

```{r}
fit_school <- lm(ATTENDANCE_RATE ~ `% Free and Reduced`, data = combined_all)
fit_school_summary <- summary(fit_school)
intercept_school <- fit_school_summary$coefficients[1, 1] 
coefficient_school <- fit_school_summary$coefficients[2, 1]  
rsquared_school <- fit_school_summary$r.squared 
school_results <- data.frame(Parameter = c("Intercept", "Coefficient for % Free/Reduced Lunch", "R-squared Value"),
                             Estimate = c(intercept_school, coefficient_school, rsquared_school))
knitr::kable(school_results, caption = "Linear Model Results", digits = 3)
```

The intercept estimate predicts that a school district with 0% of students eligible for free/reduced-price lunch will have an attendance rate of 96.2%. 

The coefficient estimates that for every 1% increase in the proportion of students eligible for free/reduced-price lunch within a school district, the attendance rate decreases by 0.061%. This result supports the hypothesis that higher levels of socioeconomic disadvantaged students, as measured by lunch eligibility, are associated with lower student attendance rates.  

The R-squared value suggests that about 28.3% of the variation in student attendance rates can be explained by the percentage of students eligible for free/reduced-price lunch. This suggests that while free/reduced lunch eligibility is a meaningful predictor, there are other factors at play that also affect attendance rates.
